Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (DLPFC with TMM normlaisation)

### 2016-09-22 code/Rmd/CMC_DLPFC_StarAligned_TMM.Rmd
Modified DLPFC processing to include custom design

---
title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (DLPFC with TMM normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

@@

library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

@@

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Ethnicity", "Age of Death", "PMI (in hours)", 
                  "pH", "Dx", "CDR", "CauseOfDeath", "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                  "DLPFC_RNA_isolation: RNA Isolation Batch", "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?",
                  "DLPFC_RNA_report: Library Batch")

@@

METADATA = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, PMI__in_hours_, pH, Dx, CDR, CauseOfDeath,
                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RNA_Isolation_Batch,
                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Library_Batch) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                IsolationBatch = DLPFC_RNA_isolation__RNA_Isolation_Batch,
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                LibraryBatch = DLPFC_RNA_report__Library_Batch) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch),
                IsolationBatch = as.factor(IsolationBatch))
```

### Data preprocessing
```{r preprocess.data, echo=TRUE}
# Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
METADATA <- METADATA %>%
  filter(SampleID %in% colnames(COUNT)) %>%
  filter(IsolationExclude == 0, ReportExclude == 0) %>%
  filter(!is.na(PMI)) %>%
  filter(!is.na(Gender)) %>%
  filter(!is.na(Ethnicity)) %>%
  filter(!is.na(RIN)) %>%
  filter(!(SampleID %in% c('MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_60', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_361', 
                           'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_389', 'PENN_RNA_PFC_56', 'PENN_RNA_PFC_59'))) %>%
  as.data.frame()
```

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Institution", "Gender", "Ethnicity", "CDR", "CauseOfDeath", "IsolationBatch", "LibraryBatch", "Dx")
ContCovariates <- c("Age_of_Death", "PMI", "pH", "RIN")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES <- data.frame(lapply(COVARIATES,function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'\\+','')})}))
rownames(COVARIATES) <- METADATA$SampleID

@@

# pH
p[[4]] = ggplot(COVARIATES, aes(x = Dx, y = pH)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('pH') + my.theme

multiplot(plotlist = p, cols = 2)
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

@@

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "Dx" is chosen as the primary variable of interest
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'RIN';#'Dx'

@@

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

# Add primary variable to the list (in case not chosen by iterative method)
postAdjustCovars = unique(c('Dx', postAdjustCovars))

@@

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "jul2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")

@@

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=8, fig.width=16}
# Fit contrast
contrast = makeContrasts(contrasts=c("DxControl-DxAFF", "DxControl-DxBP",
                                     "DxControl-DxSCZ", "DxAFF-DxBP",
                                     "DxAFF-DxSCZ", "DxBP-DxSCZ"),
                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:6, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
    rownameToFirstColumn('ensembl_gene_id') %>%
    left_join(Ensemble2HGNC) %>%
    dplyr::select(ensembl_gene_id, hgnc_symbol, logFC, AveExpr, t, P.Value, adj.P.Val, B)
}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC)
names(DE) = colnames(contrast)

>>>>

---
title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (DLPFC with TMM normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

@@

library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)

library(synapseClient)
library(knitr)
library(githubr)

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

@@

# Get ancestry vector calculated using gemtools
ANCESTRY_ID <- 'syn2511399'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
ANCESTRY <- read.table(synGet(ANCESTRY_ID)@filePath, fill = T, header = T, sep = '\t') %>%
  plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'GenotypingSampleID'))

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
                          "Dx", "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
                          'DNA_genotyping_report: Genotyping Sample_ID')

@@

METADATA = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, DLPFC_RNA_isolation__Exclude_,
                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_,
                DLPFC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                LibraryBatch = DLPFC_RNA_report__Library_Batch,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
  left_join(ANCESTRY)

# Replace AFF and BP samples as other
levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
```

### Data preprocessing
```{r preprocess.data}
# Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
writeLines('Following counts are missing any metadata')
writeLines(paste(setdiff(colnames(COUNT), levels(METADATA$SampleID)), collapse = ', '))
METADATA <- METADATA %>% filter(SampleID %in% colnames(COUNT))

writeLines('Following samples are marked exclude')
writeLines(paste(METADATA$SampleID[METADATA$IsolationExclude == 1 | METADATA$ReportExclude == 1], collapse = ', '))
METADATA <- METADATA  %>% filter(IsolationExclude == 0, ReportExclude == 0)

writeLines('Following samples are missing PMI information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$PMI)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(PMI))

writeLines('Following samples are missing gender information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$Gender)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(Gender))

writeLines('Following samples are missing RIN information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$RIN)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(RIN))

writeLines('Following samples are missing EV information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$EV.1)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(METADATA$EV.1))

writeLines('Following samples are tagged as outliers')
METADATA <- METADATA  %>% filter(!(SampleID %in% c('MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_60', 
                                                   'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_361', 
                                                   'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_389', 
                                                   'PENN_RNA_PFC_56', 'PENN_RNA_PFC_59'))) %>%
  as.data.frame()
```

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Institution", "Gender", "LibraryBatch", "Dx")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
rownames(COVARIATES) <- METADATA$SampleID

@@

multiplot(plotlist = p, cols = 3)
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to NA
log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix[log.mat] = NA

# TMM normalisation
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix)

@@

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "Dx" is chosen as the primary variable of interest
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'RIN';

@@

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
```

### Normlaisation (Custom design)
To be consistent with DLPFC Nature Neuroscience article we performed a custom adjustment with Dx, RIN, RIN2, Age of Death, PMI, Library Batch, Institution, Gender, EV1, EV2, EV3, EV4, EV5
```{r custom.design}
# Add primary variable to the list (in case not chosen by iterative method)
postAdjustCovars = c('Dx', c(FactorCovariates, ContCovariates)) %>% unique()

@@

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "mar2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")

@@

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=8, fig.width=16}
# Fit contrast
contrast = makeContrasts(contrasts=c("DxControl-DxOther",
                                     "DxControl-DxSCZ",
                                     "DxOther-DxSCZ"),
                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:3, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
    rownameToFirstColumn('ensembl_gene_id') %>%
    left_join(Ensemble2HGNC) %>%
    dplyr::select(ensembl_gene_id, hgnc_symbol, logFC, AveExpr, t, P.Value, adj.P.Val, B)
}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC)
names(DE) = colnames(contrast)
